---
title: 'Some everyday data tasks: a few hints with R'
author: "Andrea Onofri"
date: 2019-03-27
tags: R
categories: R
archives: ["2019"]
---


We all work with data frames and it is important that we know how we can reshape them, as necessary to meet our needs. I think that there are, at least, four routine tasks that we need to be able to accomplish:

1. subsetting
2. sorting
3. casting
4. melting

Obviously, there is a wide array of possibilities; I'll just mention a few, which I regularly use.

#Subsetting the data

Subsetting means selecting the records (rows) or the variables (columns) which satisfy certain criteria. Let's take the 'students.csv' dataset, which is available on one of my repositories. It is a database of student's marks in a series of exams for different subjects.

```{r}
students <- read.csv("https://www.casaonofri.it/_datasets/students.csv", header=T)
head(students)
```

Let's say that we want a new dataset, which contains only the students with marks higher than 28.

```{r}
subData <- subset(students, Mark >= 28)
head(subData)
```

Let's make it more difficult and extract the records were mark is ranging from 26 to 28 (margins included. Look at the AND clause):

```{r}
subData <- subset(students, Mark <= 28 & Mark >=26)
head(subData)
```

Now we are interested in those students who got a mark ranging from 26 to 28 in MATHS (please note the equality relationship written as '=='):

```{r}
subData <- subset(students, Mark <= 28 & Mark >=26 & 
                    Subject == "MATHS")
head(subData)
```

Lets' look for good students either in MATHS or in CHEMISTRY (OR clause; note the '|' operator):

```{r}
subData <- subset(students, Mark <= 28 & Mark >=26 & 
                    Subject == "MATHS" | 
                    Subject == "CHEMISTRY")
head(subData)
```

We can also select columns; for example we may want to display only the 'Subject', 'Mark' and 'HighSchool' columns:

```{r}
subData <- subset(students, Mark <= 28 & Mark >=26 & 
                    Subject == "MATHS" | 
                    Subject == "CHEMISTRY",
                  select = c(Subject, Mark, HighSchool))
head(subData)
```

We can as well drop the unwanted columns:


```{r}
subData <- subset(students, Mark <= 28 & Mark >=26 & 
                    Subject == "MATHS" | 
                    Subject == "CHEMISTRY",
                  select = c(-Id, 
                             -Date,
                             -Year))
head(subData)
```


Using the function 'subset()' is very easy. However, we might have higher flexibility if we subset by using indices. We already know that the notation 'dataframe[i,j]' returns the element in the i-th row and j-th column in a data frame. We can of course replace i and j with some subsetting rules. For example, taking the exams where the mark is comprised between 25 and 29 is done as follows:

```{r}
subData <- students[(students$Mark <= 29 & students$Mark >=25),]
head(subData)
```

This is useful to quickly edit the data. For example, if we want to replace all marks from 25 to 29 with NAs (Not Available), we can simply do:

```{r}
subData <- students
subData[(subData$Mark <= 29 & subData$Mark >=25), "Mark"] <- NA
head(subData)
```

Please note that I created a new dataset to make the replacement, in order not to modify the original dataset. Of course, I can use the 'is.na()' function to find the missing data and edit them.

```{r}
subData[is.na(subData$Mark) == T, "Mark"] <- 0 
head(subData)
```

#Sorting the data

Sorting is very much like subsetting by indexing. I just need to use the 'order' function. For example, let's sort the students dataset by mark:

```{r}
sortedData <- students[order(students$Mark), ]
head(sortedData)
```

We can also sort by decreasing order:

```{r}
sortedData <- students[order(-students$Mark), ]
head(sortedData)
```

We can obviously use multiple keys. For example, let's sort by subject within marks:

```{r}
sortedData <- students[order(-students$Mark, students$Subject), ]
head(sortedData)
```

If I want to sort in decreasing order on a character variable (such as Subject), I need to use the helper function 'xtfrm()':

```{r}
sortedData <- students[order(-students$Mark, -xtfrm(students$Subject)), ]
head(sortedData)
```


#Casting the data

When we have a dataset in the LONG format, we might be interested in reshaping it in the WIDE format. This is the same as what the 'pivot table' function in Excel does. For example, take the 'rimsulfuron.csv' dataset in my repository. This contains the results of a randomised block experiment, where we have 16 herbicides in four blocks. The dataset is in the LONG format, with one row per plot.

```{r}
rimsulfuron <- read.csv("https://www.casaonofri.it/_datasets/rimsulfuron.csv", header=T)
head(rimsulfuron)
```

Lets'put this data frame in the WIDE format, with one row per herbicide and one column per block. To do so, I usually use the 'cast()' function in the 'reshape' package.

```{r}
library(reshape)
castData <- cast(Herbicide ~ Block, data = rimsulfuron,
     value = "Yield")
head(castData)
```

Other similar functions are available within the 'reshape2' and 'tidyr' packages.

#Melting the data

In this case we do the reverse: we transform the dataset from WIDE to LONG format. For this task, I like the 'melt()' function in the 'reshape2' package, which requires a data frame as input. I would like to use the 'castData' object which we have just created by using the 'cast()' function above. Unfortunately, this object has a 'cast_df' class. Therefore, in order to avoid weird results, I need to change 'castData' into a data frame, by using the 'as.data.frame()' function. 

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(reshape2)
castData <- as.data.frame(castData)
mdati <- melt(castData,
              variable.name = "Block",
              value.name = "Yield",
              id=c("Herbicide"))

head(mdati)
```


Have a nice work with these functions!



