---
title: "Back-transformations with emmeans()"
author: "Andrea Onofri"
date: 2023-11-30
slug: 'Marginal means'
categories:
  - R
  - 'Linear_models'
tags:
  - R
  - 'linear_models'
archives:
  - 2023
---

```{r setup, cache = F, echo = F}
#Put at the beginning
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
knitr::knit_hooks$set(document = function(x){ 
  gsub("```\n*```r*\n*", "", x) 
})
```


I am one of those old guys who still uses the stabilising transformations, when the data do not conform to the basic assumptions for ANOVA. Indeed, apart from counts and proportions, where GLMs can be very useful, I have not yet found a simple way to deal with heteroscedasticity for continuous variables, such as yield, weight, height and so on. Yes, I know, Generalised Least Squares (GLS) can be useful to fit heteroscedastic models, but I would argue that stabilising transformations are, conceptually, very much simpler and they can be easily thought to PhD students and practitioners, with only a basic level of knowledge about statistics.

In the recent past, the package `emmeans` gave me a big boost for its useful way to handle back-transformations. For example, in the box below, I have performed an ANOVA on log-transformed data and retrieved the back-transformed means on the original count scale.

```{r}
library(emmeans)
fileName <- "http://www.casaonofri.it/_datasets/insects.csv"
dataset <- read.csv(fileName)
head(dataset)
model <- lm(log(Count) ~ Insecticide, data = dataset)
emmeans(model, ~Insecticide, type = "response")
```

It is very simple: `emmeans` auto-detects the transformation function (which is made inside the model specification) and automatically produces the back-transformation, when this is requested by using the '`type = "response"`' argument (we can also use the argument '`regrid = "response"`', with slight differences that I will discuss in a future post).

Unfortunately, not all transformations are auto-detected; for example, let's consider the dataset 'Hours_to_failure.csv', where we have the time-to-failure (in hours) for a device, as affected by the operating temperature. If we regard the temperature as a factor, we can fit an ANOVA model; a check with the `boxcox()` function in the MASS package suggests that this dataset might require a stabilising transformation into the inverse value. 

```{r}
library(MASS)
library(emmeans)
fileName <- "http://www.casaonofri.it/_datasets/Hours_to_failure.csv"
dataset <- read.csv(fileName)
dataset$Temp <- factor(dataset$Temp)
head(dataset)
model <- lm(Hours_to_failure ~ Temp, data = dataset)
tp <- boxcox(model)
```

As we see below, the inverse transformation is not auto-detected.

```{r}
model2 <- lm(I(1/Hours_to_failure) ~ Temp, data = dataset)
emmeans(model2, ~ Temp, type = "response")
```

In this situation, an alternative approach must be used. The transformation can be made prior to fitting the model; next, we need to update the 'reference grid' for the model, specifying what type of transformation we have made (`tran = "inverse"`). Finally, we can pass the updated grid into the `emmeans()` function. And ... the back-transformation is served!

```{r}
dataset$invHours <- 1/dataset$Hours_to_failure
model3 <- lm(invHours ~ Temp, data = dataset)
updGrid <- update(ref_grid(model3), tran = "inverse")
emmeans(updGrid, ~Temp, type = "response")
```

I can use this method with several functions, such as: "identity", "1/mu^2", "inverse", "reciprocal", "log10", "log2", "asin.sqrt", and "asinh.sqrt".

Sometimes, I need extra-flexibility. For example, if we look at 'boxcox' graph above, we see that the inverse transformation, although acceptable (the value $-1$ is within the confidence limits for $\lambda$) is not the best one. Indeed, the maximum likelihood value is -0.62:

```{r}
tp$x[which.max(tp$y)]
```

Therefore, we might like to use the following transformation, that is within the 'boxcox' family:

$$W = \frac{Y^{-0.62} - 1}{-0.62}$$

This type of transformation is not manageable with the previous code and we need to use the `make.tran()` function, specifying the value of $\lambda$ (`alpha = -0.62`) and the value of the displacement parameter (`beta = 1`).

```{r}
dataset$bcHours <- (dataset$Hours_to_failure^(-0.62) - 1)/(-0.62)
model4 <- lm(bcHours ~ Temp, data = dataset)
updGrid <- update(ref_grid(model4), 
                  tran = make.tran("boxcox", alpha = -0.62,
                                   beta = 1))
emmeans(updGrid, ~Temp, type = "response")
```

The function `make.tran()` can be used to specify several other transformation functions, such as the angular transformation that is often used for percentages and proportions. You can get a full list by searching help ('?make.tran') from inside R.

In conclusion, stabilising transformations, in spite of their age, can still be useful to fit heteroscedastic models; do not underrate them, just because they are no longer in fashion!

Thanks for reading!

Prof. Andrea Onofri   
Department of Agricultural, Food and Environmental Sciences    
University of Perugia (Italy)    
Send comments to: [andrea.onofri@unipg.it](mailto:andrea.onofri@unipg.it)

<a href="https://twitter.com/onofriandreapg?ref_src=twsrc%5Etfw" class="twitter-follow-button" data-show-count="false">Follow @onofriandreapg</a><script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>


